---
title: Who can best predict the result of the 2018 World Cup?
---


# Initialization

First we load some of the packages that are needed and set some global parameters. 

```{r message=FALSE}
library("dplyr")
library("magrittr")
library("ggplot2")

normalgoals <- 2.75  # The average number of goals scored in a world cup match
```

Now we include two datasets that are needed in the following. 

The first `team_data` contains information about the teams in the world cup. You are welcome to add more columns to this data frame but things might get messed up if you remove any of the ones listed below. If you want to add extra information about each team to use when determining the outcome of a match then add it here.

The `group_match_data` data frame contains information about the game plan. Which teams are playing and when. This should not really be modified as it should represent the actual tournament.

```{r}
## This data frams contains information about the teams.
## You are free to add information here that you can use when determining winners
team_data <- tibble(
  number = 1:32,
  name = c("Egypt","Russia","Saudi Arabia","Uruguay",
           "Iran","Morocco","Portugal","Spain",
           "Australia","Denmark","France","Peru",
           "Argentina","Croatia","Iceland","Nigeria",
           "Brazil","Costa Rica","Switzerland","Serbia",
           "Germany","South Korea","Mexico","Sweden",
           "Belgium","England","Panama","Tunisia",
           "Colombia","Japan","Poland","Senegal"),
  group = rep(LETTERS[1:8], each=4),
  rating = c(151, 41, 1001, 34,
             501, 501, 26, 7,
             301, 101, 7.5, 201,
             10, 34, 201, 201,
             5, 501, 101, 201,
             5.5, 751, 101, 151,
             12, 19, 1001, 751,
             41, 301, 51, 201),
  elo = c(1646, 1685, 1582, 1890, # From https://www.eloratings.net/, May 12th 2018
          1793, 1711, 1975, 2048,
          1714, 1843, 1984, 1906,
          1985, 1853, 1787, 1699,
          2131, 1745, 1879, 1770,
          2092, 1746, 1859, 1796,
          1931, 1941, 1669, 1649,
          1935, 1693, 1831, 1747)
)

group_match_data <- read.csv(text=
"team1,team2,date,goals1,goals2
Russia,Saudi Arabia,14/06/2018,,
Egypt,Uruguay,15/06/2018,,
Morocco,Iran,15/06/2018,,
Portugal,Spain,15/06/2018,,
France,Australia,16/06/2018,,
Argentina,Iceland,16/06/2018,,
Peru,Denmark,16/06/2018,,
Croatia,Nigeria,16/06/2018,,
Costa Rica,Serbia,17/06/2018,,
Germany,Mexico,17/06/2018,,
Brazil,Switzerland,17/06/2018,,
Sweden,South Korea,18/06/2018,,
Belgium,Panama,18/06/2018,,
Tunisia,England,18/06/2018,,
Colombia,Japan,19/06/2018,,
Poland,Senegal,19/06/2018,,
Russia,Egypt,19/06/2018,,
Portugal,Morocco,20/06/2018,,
Uruguay,Saudi Arabia,20/06/2018,,
Iran,Spain,20/06/2018,,
Denmark,Australia,21/06/2018,,
France,Peru,21/06/2018,,
Argentina,Croatia,21/06/2018,,
Brazil,Costa Rica,22/06/2018,,
Nigeria,Iceland,22/06/2018,,
Serbia,Switzerland,22/06/2018,,
Belgium,Tunisia,23/06/2018,,
South Korea,Mexico,23/06/2018,,
Germany,Sweden,23/06/2018,,
England,Panama,24/06/2018,,
Japan,Senegal,24/06/2018,,
Poland,Colombia,24/06/2018,,
Saudi Arabia,Egypt,25/06/2018,,
Uruguay,Russia,25/06/2018,,
Iran,Portugal,25/06/2018,,
Spain,Morocco,25/06/2018,,
Australia,Peru,26/06/2018,,
Denmark,France,26/06/2018,,
Nigeria,Argentina,26/06/2018,,
Iceland,Croatia,26/06/2018,,
Mexico,Sweden,27/06/2018,,
South Korea,Germany,27/06/2018,,
Serbia,Brazil,27/06/2018,,
Switzerland,Costa Rica,27/06/2018,,
England,Belgium,28/06/2018,,
Senegal,Colombia,28/06/2018,,
Panama,Tunisia,28/06/2018,,
Japan,Poland,28/06/2018,,
",header=TRUE)
```



## Skellam distribution 

All of this section is irrelevant and it is used to compute match results using a Skellam distribution (difference of Poissons) when there is a restriction on the parameters based on the skill levels of the teams. Disregard this at first

```{r skellam}
dskellam <- function(x, mu1, mu2) {
  return(exp(-(mu1+mu2))*(mu1/mu2)^(x/2)*besselI(2*sqrt(mu1*mu2),nu=x)
         )
}


eta <- 2.75
beta1 <- seq(0.05, eta-0.05, 0.05)


skellam <- rep(0, length(beta1))
counter <- 1
for (i in beta1) {
  
  # Compute probability that team 1 wins
  skellam[counter] <- sum(dskellam(1:12, i, eta-i)) / ( sum(dskellam(seq(-10,10,1), i, eta-i)) - dskellam(0, i, eta-i) )
  counter <- counter+1
}

skellam <- data.frame(beta=beta1, prob=skellam)


FindParameter <- function(prob) {
      sapply(prob, function(i) {
                 if (i<.009) {
                     return (.1)
                 }
                 if (i>.995) {
                       return (eta-.05)
                 }
               return(min(skellam$beta[skellam$prob>i]))
           })
}

```


# Play game

This function is the **most important** to modify as it is the one that determines the outcome (a vector of length 2) of goals scored for each team. **This** is the function to modify and where you should add your own model for predicting the outcome of a single match. It can use the `team_data` from above and `team1` and `team2` include vectors of the team numbers that are playing against each other. `play_game` should return a matrix with dimensions `length(team1)` x `2` with the goals scored by each team.

```{r}
# Input: First three arguments are fixed. Default values are given
# Returns: a matrix with length(team1) rows 
# and 2 columns with goals
play_game <- function(team_data, team1, team2, musthavewinner=FALSE, ...) {
  # Sanity checks
  if (length(team1) != length(team2))
    stop("Lengths of team should be the same")
  
  if (any(team1==team2))
    stop("A team cannot play against itself")

  ## Simplest version. 
  ## All teams are equal
  result <- cbind(rpois(length(team1), lambda=normalgoals/2), 
                  rpois(length(team1), lambda=normalgoals/2))

  ## Skellam distribution
  ## Uncomment the code below to use the skellam model
  p1 <- .91/team_data$rating[team1]
  p2 <- .91/team_data$rating[team2]
  prob <- p1 / (p1 + p2)
  lambdaA <- FindParameter(prob)
  Agoals <- rpois(length(prob), lambdaA)
  Bgoals <- rpois(length(prob), normalgoals-lambdaA)
  result <- cbind(Agoals, Bgoals)
  
  ## ELO version (no update here). Using sapply here instead of
  ## vectorization in case the elo ranking should be updated after each match.
  # Uncomment below to use the ELO model
#  result <- t(sapply(seq_len(length(team1)), function(i) {
#                          AWinProb <- 1/(1 + 10^((team_data$elo[team2[i]] - team_data$elo[team1[i]])/400))
#                          myres <- rbinom(1, size=1, prob=AWinProb)
#                              fakegoals <- c(1,0)  
#                              if (myres==0)
#                                fakegoals <- c(0,1)
#                              fakegoals
#                             }))

  # If we MUST have a winner then one simple trick is to add a random goal 
  # to one of the two teams that have the same score. Penalty goals seem rather 
  # random anyway
  if (musthavewinner) {
    result[result[,1]==result[,2],1] + 2*rbinom(sum(result[,1]==result[,2]), size=1, prob=.5) - 1
    
  }
  result
}
```


# Helper functions

The following chunk contains helper functions to run the simulations and find the winners at various stages.

```{r}
find_group_winners <- function(team_data, group_match_data) {

  ## Create a copy of the the matches that we can fill out
  group_match_results <- group_match_data

  ## Simulate each match that hasn't already been played  
  pick <- (!complete.cases(group_match_results[c("goals1", "goals2")]))
  group_results <- play_game(team_data, 
                             team_data$number[fmatch(group_match_data$team1[pick], team_data$name)],
                             team_data$number[fmatch(group_match_data$team2[pick], team_data$name)],
                             musthavewinner = FALSE)

  ## Now add the results (the goals) to the match resuls
  group_match_results[pick, c("goals1", "goals2")] <- group_results

  ## Compute points earned per team for each match
  group_match_results$pointsForA <- with(group_match_results, 3*(goals1>goals2)+1*(goals1==goals2))
  group_match_results$pointsForB <- with(group_match_results, 3*(goals1<goals2)+1*(goals1==goals2))


  ## Okay the casing is a bit of a mess here. I do apologize.
  team_data$points <- 
  sapply(team_data$name, function(i) { sum(group_match_results[c("pointsForA", "pointsForB")][i == group_match_data[c("team1","team2")]]) })
  team_data$goalsFore <- sapply(team_data$name, function(i) { sum(group_match_results[c("goals1", "goals2")][i == group_match_data[c("team1","team2")]]) })

  team_data$goalsAgainst <- sapply(team_data$name, function(i) { sum(group_match_results[c("goals2", "goals1")][i == group_match_data[c("team1","team2")]]) })
  
  team_data$goalsDifference <- team_data$goalsFore-team_data$goalsAgainst


  # And here we find the winners within each group
  team_data %>% 
    group_by(group) %>% 
    arrange(desc(points), desc(goalsDifference), desc(goalsFore)) %>% 
    mutate(groupRank = row_number()) %>% 
    ungroup() %>%
    arrange(group, groupRank)
}


find_knockout_winners <- function(team_data, match_data) {
  ## Get the results
  results <- play_game(team_data, match_data[,1], match_data[,2], musthavewinner=TRUE)
  ## Find the teams that won
  winners <- match_data[cbind(seq(nrow(results)), ifelse(results[,1]>results[,2], 1, 2))]
  winners
}



simulate_tournament <- function(n=10,
                                teams=team_data, 
                                group_matches=group_match_data) {
  
  
  sapply(1:n, function(matchnumber) {
    
  ## Step 1: Find the results from the group matcges
  group_results <- find_group_winners(team_data=teams, group_matches)
  
  ## Step 2: Design matches for the first part of the knockout match
  eigth_matches <- cbind(group_results$number[seq(1, 32, by=4)], group_results$number[c(6, 2, 14, 10, 22, 18, 30, 26)])
  ## and find the results
  eigth_winners <- find_knockout_winners(team_data, eigth_matches)

  ## Step 3: Design matches for the quarter finals and run them
  quarter_matches <- cbind(eigth_winners[c(1, 2, 5, 6)], eigth_winners[c(3, 4, 7, 8)])
  quarter_winners <- find_knockout_winners(team_data, quarter_matches)

  ## Step 4: Semi finals ... yada yada yada
  semi_matches <- cbind(quarter_winners[c(1,2)], quarter_winners[c(3,4)])
  semi_winners <- find_knockout_winners(team_data, semi_matches)

  ## Steps 5 and 6 Find number 1-4
  bronze_match <- matrix(quarter_winners[!quarter_winners %in% semi_winners], ncol=2)
  bronze_winner <- find_knockout_winners(team_data, bronze_match)

  final_match <- matrix(semi_winners, ncol=2)
  final_result <- find_knockout_winners(team_data, final_match)

  ## Return a vector with the teams in ranked order. 
  ## Note only the first 4 are individuals - the rest are really groups
  final_ranking <- c(final_result, # Number 1
                     final_match[!(final_match %in% final_result)], #2
                     bronze_winner, # Number 3
                     bronze_match[!(bronze_match %in% bronze_winner)], #4
                     quarter_matches[!(quarter_matches %in% quarter_winners)], # 5-8
                     eigth_matches[!(eigth_matches %in% eigth_winners)], # 9-16
                     seq(32)[!(seq(32) %in% eigth_matches)]
                   )
 
  final_ranking 
  })
}
```

# Run the tournament

Now we can simulate the tournament
```{r}
result <- simulate_tournament(100000)  # Run 10000 world cups

```

Then we can get a list of winners (and the probabilities of winning the full tournament:

```{r}
winner <- table(result[1,])
names(winner) <- team_data$name[match(names(winner), team_data$number)]
sort(winner/sum(winner)*100, decreasing=TRUE)
```

To compute the log-loss we first need a 32x32 matrix with rankings (rows) and countries (columns). Note that each row and column add to 1

```{r}
# Ranks are rows, columns are countries
prediction_matrix <- sapply(1:32, function(i) {rowMeans(result==i)})
prediction_matrix
```


The `logloss_score` function accepts a vector of length 32 of the true rankings (the `trueranking` argument) which we can only fill in after July 15th when the world cup is over. If team 1 won, team 2 became second etc, then `trueranking=1:32`. The argument `matrix_prediction` is the prediction matrix computed above with the probabilities for each rank and each country. 

The function automatically groups the ranks into fewer groups since we cannot differentiate between ranks 5-8, 9-16, and 17-32.

```{r}
logloss_score <- function(trueranking, matrix_prediction, 
                          weights=c(1, 1, 2, 2, 4, 4, 4, 4,
                                    8,8,8,8,8,8,8,8,
                                    rep(16, 16)),
                          threshold=1/64) {
  ## Minimal sanity check
  stopifnot(length(trueranking)==32)
  
  rank_group <- c(1, 2, 3, 4, rep(5, 4), rep(6, 8), rep(7, 16))
  rank_group_size <- c(1, 1, 1, 1, 4, 8, 16)
  
  comb <- rbind(c(1, rep(0, 31)),
        c(0, 1, rep(0, 30)),
        c(0, 0, 1, rep(0, 29)),
        c(0, 0, 0, 1, rep(0, 28)),
        c(rep(0, 4), rep(1, 4), rep(0, 24)),
        c(rep(0, 8), rep(1, 8), rep(0, 16)),
        c(rep(0:1, times=c(16, 16)))
        )

  condensed_prediction_matrix <- comb %*% matrix_prediction
  
  print(cbind(rank_group, trueranking,
    log(pmin(pmax(condensed_prediction_matrix[cbind(rank_group, trueranking)], 
                    threshold), 
               1-threshold))))
  
  sum(log(pmin(pmax(condensed_prediction_matrix[cbind(rank_group, trueranking)], 
                    threshold), 
               1-threshold))/weights)
  
}

## Here's an example of calculating with weighted loss
logloss_score(sample(1:32), prediction_matrix)
logloss_score(c(17, 8, 21, 1:7, 9:16, 18:20, 22:32), prediction_matrix)

```


# Enjoy!
